% -----------------------------------------------------------------------------
% -*-TeX-*- -*-Hard-*- Smart Wrapping
% -----------------------------------------------------------------------------
% Thesis Acknowledgements -----------------------------------------------------

\prefacesection{Acknowledgements}
\def\baselinestretch{1.0}
\setlinespacing{1.15}


I would like to thank Heydar Radjavi, my supervisor, for his many suggestions
and constant support during this research. I am also thankful to Matja\v{z}
Omladi\v{c} for his guidance through the early years of chaos and confusion.

Professor Y.A.~Abramovich expressed his interest in my work and supplied me
with the preprints of some of his recent joint work with C.D.~Aliprantis and
O.~Burkinshaw, which gave me a better perspective on my own results. Heinz
Bauschke shared with me his knowledge of convex analysis and provided many
useful references and friendly encouragement.

I had the pleasure of meeting the Killam trustees. They are wonderful people
and their support makes research like this possible. The {\em Izaak Walton
Killam Memorial Scholarship}, which was awarded to me for the period
1991--1994, was crucial to the successful completion of this project.

I should also mention that my graduate studies in Canada were supported in
part by the Research Council of Slovenia.

Of course, I am grateful to my parents for their patience and {\em love}.
Without them this work would never have come into existence (literally).

\medskip

Finally, I wish to thank the following: \, Francisco (for his friendship);
Adriana (for changing my life from worse to bad); Maks and Lili (for the
painful introduction to the art of paragliding); Karen, Kelly, Linda, Rick,
Maja, Paul, Phil, Michelle, Peter, Wendy, Scott, Eve, George, Andrew, Rob ...
(for all the good and bad times we had together); Stephen and Bill (for
special effects); Ludwig van Beethoven and The Rolling Stones (and they know
why); {\em and } my brother (because he asked me to).

\bigskip\medskip

\noindent
Halifax, Nova Scotia \hfill Aleksander Simoni\v{c}\\
December 7, 1994

% -----------------------------------------------------------------------------
